---
title: "R for Data Science"
author: "Charles Kraemer"
date: "2023-04-10"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```


```{r data, include=FALSE, echo=FALSE}
library(tidyverse)
library(ggpubr)
library(plotly)
library(kableExtra)
library(broom)

data = read_csv("gapminder_clean.csv")
```

# CO2 Emissions and GDP per Capita in 1962

```{r 1962, echo=FALSE}
data %>%
  filter(Year == 1962) %>%
  ggplot(aes(x = `CO2 emissions (metric tons per capita)`, y = gdpPercap)) +
  geom_point() +
  stat_cor(method = "pearson") + 
  labs(title="GDP with Respect to CO2 Emissions in Tons (per Capita) 1962")
```

# Year of Strongest Correlation between CO2 Emissions and GDP per Capita

```{r findMaxCorrYear, echo=FALSE}
maxCorrYear <- data %>%
  group_by(Year) %>%
  summarise(emissions_gdp_cor=cor(`CO2 emissions (metric tons per capita)`, gdpPercap, use="complete.obs")) %>%
  slice_max(emissions_gdp_cor) %>%
  select(Year)
kbl(maxCorrYear)
```

# Interactive Scatterplot Comparing CO2 Emissions and GDP per Capita in Year of Strongest Correlation

```{r withinMaxCorrYear, echo=FALSE}
gg <- data %>%
  filter(Year == maxCorrYear[[1]]) %>%
  ggplot(aes(x = `CO2 emissions (metric tons per capita)`, y = gdpPercap)) +
  geom_point(aes(size=pop, color=continent)) +
  labs(title="GDP with Respect to CO2 Emission in Tons (per Capita)")
p <- ggplotly(gg)
p
```

# Relationship between Continent and Energy Use

Below we will investigate the relationship between continent and energy use (kg of oil per capita) spanning the entirety of the dataset.  Given that continent is a nonbinary, categorical variable and energy use is a quantitative variable, we will start with an ANOVA test for statistical significance.
```{r aov, echo=FALSE}
library(multcomp)

withCountry <- data[!is.na(data$continent),]
res_aov <- aov(`Energy use (kg of oil equivalent per capita)` ~ continent,
                   data=withCountry)
summary(res_aov)
```
Given that the p-value is below 0.05, at a standard confidence level of 95%, we have determined that there is some statistically significant relationship between continent and energy use in this dataset.  For more specific understanding of that relationship, we will now use Tukey HSD testing for pairwise comparisons of energy use between each two countries.
```{r tukey, echo=FALSE}
t <- TukeyHSD(res_aov, conf.level = .95)
kbl(t$continent) %>%
  kable_styling(full_width = T)
```
As seen above, there are two pairs with a p-value > 0.05: Asia-Americas and Europe-Oceania.  This observation supports a three-way clustering of energy use by continent in which one cluster consists solely of Africa, the other consists of Asia and America, and the other consists of Europe and Oceania.  Within-cluster similarity is illustrated by the below boxplot.
```{r boxp, echo=FALSE}
withCountry %>%
  ggplot(aes(x = continent, y = `Energy use (kg of oil equivalent per capita)`)) +
  geom_boxplot() + 
  labs(title="Energy use with Respect to Continent")
```

# Imports of Goods and Services as % of GDP in Europe and Asia after 1990

Below we will determine whether there is a significant difference between Europe and Asia with respect to import goods and services as a % of GDP.  Given that we are using a binary subset of the categorical continent variable for prediction and that import goods and services as a % of GDP is a quantitative variable, t-testing will be used.

```{r euro-asia-imports-analysis, echo=FALSE}
sol <- data %>%
  filter(Year >= 1990 & (continent == "Asia" | continent == "Europe")) %>%
  summarise(pval = t.test(`Imports of goods and services (% of GDP)` ~ continent, var.equal = TRUE)$p.value)
kbl(sol)
```

Given that the p-value is above 0.05, there is not a significant difference between Europe and Asia with respect to import goods and services as a % of GDP.

```{r euro-asia-imports-visual, echo=FALSE}
data %>%
  filter(Year >= 1990 & (continent == "Asia" | continent == "Europe")) %>%
  ggplot(aes(x = continent, y = `Imports of goods and services (% of GDP)`)) +
  geom_boxplot() +
  labs(title="Import Goods and Services as % of GDP")
```

The above boxplot provides a visual intuition for the lack of a significant difference between Europe and Asia with respect to import goods and services as a % of GDP. The medians for the two countries are roughly the same, and the lower and upper quartiles for both countries are both in roughly the range [25, 60], which covers less than a quarter of the full range of the data.

# Country with Highest Average Population Density over All Years
```{r country-pop, echo=FALSE}
library(dbplyr)
sol <- data %>%
  group_by(`Country Name`) %>%
  summarise(avg_pop_density = mean(`Population density (people per sq. km of land area)`)) %>%
  slice_max(avg_pop_density) %>%
  dplyr::select(`Country Name`)
kbl(sol)
```

# Country with Highest Increase in Life Expectancy from 1962 to 2007
```{r life-expec, echo=FALSE}
sol <- data %>%
  filter(Year == 1962 | Year == 2007) %>%
  group_by(`Country Name`) %>%
  summarise(life_change = `Life expectancy at birth, total (years)`[2] - `Life expectancy at birth, total (years)`[1]) %>%
  slice_max(life_change) %>%
  dplyr::select(`Country Name`)
kbl(sol)
```
